---
  output:
  html_document:
    keep_md: true
    toc: true
    toc_float: true
    exclude: true
---
  
# dplyr and tidyr
  
dplyr introduces a consistent grammar of data manipulation for R. 
The significance of this should not be underestimated. 
It makes the standard data manipulation that all data analysts need to do so much easier. 
A particularly nice feature is the ability to connect a series of commands in the same way as pipes in Unix systems. 
Rather than the vertical bar `|`, the symbol is `%>%`.

As well as dplyr, Hadley has introduced another new package for reshaping: tidyr. 
Together with ggplot2, lubridate, stringr and others they form the 'tidyverse'.

* TOC
{:toc}

## The basics

dplyr has, as described in the introductory documentation, "five basic data manipulation verbs that work on a single table: filter(), arrange(), select(), mutate() and summarise()."

## Creating new variables

Uses `mutate()`

### Detecting strings

I have a vector in a dataframe and I wish to determine whether I can detect any one of a series of strings. 
I would normally use str_detect for this, but it doesn't work without a little trickery. 
The `paste(something, collapse = "|")` is the important bit here. 

```r
test.data <- data.frame(item = c("Apple", "Bear", "Orange", "Pear", "Two Apples"))
fruit <- c("Apple", "Orange", "Pear")
test.data
item
1 Apple
2 Bear
3 Orange
4 Pear
5 Two Apples

test.data <- test.data %>% 
mutate(is.fruit = str_detect(item, paste(fruit, collapse = "|")))
test.data
item          is.fruit
1 Apple         TRUE
2 Bear          FALSE
3 Orange        TRUE
4 Pear          TRUE
5 Two Apples    TRUE
```

### Replacing values with dplyr functions

Can use `recode()` or `case_when()`. 
`recode()` is a vectorised switch. 
It's important to note that the data types must be the same when using `recode()`, and that a double is not the same type as an integer, which seems obvious when typed. 
If they are not the same type, then unspecified replacements get NA values. 

```{r}
mtcars %>% 
  mutate(new_cyl = recode(cyl, `6` = 16, `8` = 200) ) %>% 
  select(cyl, new_cyl) %>% 
  tbl_df() %>%
  head()
```

`case_when()` vectorises multiple `if()` `else()` statements, and it's a lot neater than writing nested `ifelse()` statements. 
`case_when()` can be used inside a `mutate()` call, or outside. 
If inside, then `.$` needs to be used to specify the column upon which tests should be predicated. 

```{r}
mtcars %>% 
mutate(
  new_cyl = case_when(.$cyl == 4 ~ "Four", 
                      .$cyl == 6 ~ "Six", 
                      .$cyl == 8 ~ "Eight"
                      )) %>%
  select(cyl, new_cyl) %>% 
  head()
```

There's an important gotcha here though. 
One has to make sure that each possible outcome is specified, otherwise unspecified levels are replaced with `<NA>` values.
One can work around this as in the second example. 
I haven't yet tested what will happen if numeric output is required, I expect that one would need `TRUE ~ as.double(.$cyl)`.
For numeric columns, it's also worth remembering that integer data types are not the same as other numeric data types such as doubles and so `as.numeric()` may still return `<NA>` values.

```{r}
mtcars %>% 
  mutate( cyl2 = case_when(
    .$cyl == 4 ~ "four"
  )
  ) %>% 
  select(cyl, cyl2) %>% 
  head()

mtcars %>% 
  mutate( cyl2 = case_when(
    .$cyl == 4 ~ "four", 
    TRUE ~ as.character(.$cyl)
  )
  ) %>% 
  select(cyl, cyl2) %>% 
  head()
```

This also works for detecting strings in columns. 
It's a bit weird giving the `TRUE ~ 0` for the non-true cases. 
I tried `FALSE ~ 0` and that doesn't work. 

```{r}
library(stringr)
mtcars$name <- row.names(mtcars)
names_to_detect <- c("Mazda", "Datsun")

mtcars %>% 
  mutate(has_name = case_when(
    str_detect(.$name, paste(names_to_detect, collapse = "|")) == TRUE ~ 1, 
    TRUE ~ 0)
  )%>% 
  select(name, has_name) %>% head()

```

## Working on multiple columns at once

`mutate_all` or `summarise_all` will affect all variables in the dataframe in the same way. 
`mutate_at` will mutate only the specified columns

```{r}
dat <- data.frame(x = c(1,2,3), 
  y_date = c("01/01/2017", "02/01/2017", "03/01/2017"), stringsAsFactors = FALSE)
dat <- dat %>% mutate_at(vars(contains("date")), funs(lubridate::dmy))
dat
```

### Passing arguments to `summarise_all`

For example, you might want to specify removal of NA values in a sum or mean function. 
You can do this by
```{r, eval = FALSE}
data(mtcars)
summarise_all(mtcars, funs(sum(., na.rm = TRUE)))
```

## Subsetting (aka filtering)

Much less verbose than `test[test$var1 == "something",]`
```{r, eval = FALSE}
filter(test, var1 == "something")
```
       
But getting unique combinations is a little more awkward. 
One can hijack `n_distinct()` introduced in dplyr to achieve this.

```{r}
test <- data.frame(id = c(1,1,1,2,2),
org = c("apple", "apple", "bear", "orange", "pear"),
test = c("S", "R","S", "R", "S"))
test

out <- test %>% group_by(id, org) %>% 
  filter(row_number() == 1)
out
```

## Summarising data

Uses the `summarise()` function. `n()` is a function introduced in dplyr and produces a count of the number of rows in the data set. 
                   
```{r}
library(binom)

test <- data.frame(group = c(rep("A", 10), rep("B", 10)),
outcome = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0))
head(test)

table(test$group, test$outcome, dnn = c("Group", "Outcome"))

out <-test %>% group_by(group) %>% 
  summarise(n = n(), successes = sum(outcome), 
            pc.succ = round(binom.confint(successes, n, methods = "exact")[[4]]*100, 2),
            pc.lci = round(binom.confint(successes, n, methods = "exact")[[5]]*100, 2),
            pc.uci = round(binom.confint(successes, n, methods = "exact")[[6]]*100, 2)
            )
out
```
                   
Outside dplyr I would write `binom.confint(x,n,methods = "exact")$mean`. 
This doesn't work in dplyr and returns the error `"Error in binom.confint(c(6928L, 5704L, 4746L, 5166L, 6779L, 6275L, 4832L,  : 
                                                                                                 invalid subscript type 'closure' "`
However, indexing the columns does work. 
Also, wrapping the `round()` and `binom.confint()` in a function would improve the readability of the process. 
                                                                                               
## Grouped operations

There may be a variable name length limit in `group_by()`.
I believe that this is now fixed. 
Certainly the below works.
```{r} 
test <- data.frame(name = rep(c("orange", "pear", "apple", "bear"), 2) , 
                   value = rnorm(8), stringsAsFactors = FALSE)
test %>%
  group_by(name) %>% summarise(mean(value))
                                                                                               test$reallyreallyreallyreallyreallylongvarname <- test$name

test %>% group_by(reallyreallyreallyreallyreallylongvarname) %>% 
  summarise(mean(value))
```

This can be resolved using `quote()`.
```{r}
group_by(test, quote(reallyreallyreallyreallyreallylongvarname)) %>% 
  summarise(mean(value))
```

### Adding a group id
Sometimes one might want to create an integer group id. 
i.e. for all rows with the same groupings, these will have the same id. 
`group_indices()` goes some of the way to resolve this by creating a vector of integers corresponding to the index numbers of the groups. 
Unfortunately, one can't use `group_indices()` in `mutate()`, it returns an error.
Therefore, one has to sort the dataframe by the grouping variables before adding in the id. 
It's important to note that the data **must** be sorted prior to the creation of `i`, otherwise the `mutate(group_id = i)` will be in the wrong sort order.

```{r}
library(dplyr)
data(mtcars)
mtcars <- mtcars %>% group_by(cyl, am) %>% arrange(cyl, am)
i <- mtcars %>% group_indices(cyl, am)
mtcars <- mtcars %>% ungroup() %>% mutate(group_id = i)
mtcars %>% select(cyl, am, group_id)
```

## Dropping and renaming variables

Variables can be renamed with `rename()`. 
The syntax is `rename(data, new.var = old.var)`. 
Note that there are no quotations marks. 
e.g. 

```{r}
rename(iris, petal_length = Petal.Length) %>% head()
```

Variables can be dropped using `select(data, -var.name)`. 
Or, if quicker, retained from a long list of variables, e.g. `select(data, var1, var2)`.
                                                                                               
```{r}
select(iris, -Sepal.Length) %>% head
```

```{r}
select(iris, Sepal.Length, Species) %>% head
```
                                                                                               
It's also possible to select columns based on text that the column name contains:
                                                                                               
```{r}
pps <- pps %>% select(p.full.id, contains("_atc"))
```
                                                                                               
thanks to: http://stackoverflow.com/questions/25923392/r-dplyr-select-columns-based-on-string
                                                                                               
## Lagged values and windows
```{r}
#  test windowed functions
require(dplyr)

test <-
data.frame(cbind(time = c(1:10), value = rep(1, 10)), stringsAsFactors = FALSE)
head(test)
```

```{r}
test <- test %>% 
  mutate(
  lag1 = lag(value),
  lag2 = lag(value, 2),
  lag3 = lag(value, 3),
  lag4 = lag(value, 4),
  rolling_4 = rowSums(.[3:6], na.rm = TRUE)
)
test
```
                                                                                        ## Wrapping dplyr in a function

Programming with dplyr has changed a lot with the release of dplyr 0.7. 
So, much of what follows is likely to be inaccurate for version 0.7 and later. 

Can be a little tricky and some of the information on the internet seems out of date. However, managed to wrap the following in a function to produce nice univariable tables. It does require a variable called obs though. This should be fixable with a little wrangling. 
                                                                                               
```{r}

niceTable <- function(data = x, ...){
  # from https://groups.google.com/d/msg/manipulatr/htt0kO9Dhds/O5eVE2cfvsoJ
  
  z <- data %>% group_by_(...) %>%
    summarise(n = sum(obs)) %>% ungroup() %>% 
    mutate(cum.pc = round((cumsum(n)/sum(n))*100,1), pc = round( (n/sum(n))*100, 2 ))
  return(z)
  }

data(mtcars)
mtcars$obs <- 1

out <- niceTable(mtcars, "vs")
out
```

Done.^[But `tabyl()` in janitor makes this nicer]
Or so I thought. To go much further with this is actually quite difficult. 
Let's try without wrapping in a function:

```{r}
require(dplyr)
mtcars <- mtcars
mtcars %>% group_by(cyl) %>% summarise(n = n(), am = sum(am)) %>%
  mutate(pc.am = round((am/n)*100, 2))
```
                                                                                         Then, to get this in a function, one needs to supply more than one variable to the data. 
                                                                                         This turns out to be rather tricky. 
                                                                                         One has to interpret the multiple variables supplied, otherwise you get all sorts of errors. 
```{r}
                                                                                               require(lazyeval) # I thought dplyr loaded this by default, but apparently not. 
                                                                                               tabFun <- function(x, y, z){
                                                                                               a <- x %>% group_by_(as.name(y)) %>% 
                                                                                               summarise_(n = interp(~length(var), var = as.name(z)), 
                                                                                               sum.z = interp(~sum(var), var = as.name(z))) %>%
                                                                                               mutate(pc.am = round((sum.z/n)*100, 2))
                                                                                               return(a)
                                                                                               }
                                                                                               
                                                                                               tabFun(mtcars, "cyl", "am")
```
Phew. 

Oh, and incidentally, trying to simplify this with n = n() returns the error `Error in n() : This function should not be called directly`. 
                                                                                         I don't know why. 

## Mutate in a function
ok me a long time to find a solution for this. 

I tinkered around with `quote()` and `substitute()` for a bit. 
I don't understand why a simple `mutate_()` doesn't work. 
I don't really see the need for `[]` indexing, but the following allows you to avoid quotes in the call. 

```{r, eval=FALSE}

age_fun <- function(data, dob, sampledate){

  var1 <- deparse(substitute(dob))
  var2 <- deparse(substitute(sampledate))
                                                                                           data <- mutate(data, age = data[, var2] - data[, var1],
                 age = as.numeric(age)/365.25)
                 
                                                                                           }
ec <- age_fun(ec, date_of_birth, sample_date)
                                                                                         #      from http://stackoverflow.com/questions/24606282/passing-data-frame-to-mutate-within-function
 ```
                                                                                               
                                                                                               ## Wrapping filter in a function
                                                                                               
                                                                                               ```r
                                                                                               require(dplyr)
                                                                                               data(mtcars)
                                                                                               mtcars$name <- row.names(mtcars)
                                                                                               
                                                                                               just_1 <- mtcars %>% filter(name == "Mazda RX4")
                                                                                               just_1
                                                                                               
                                                                                               mpg cyl disp  hp drat   wt  qsec vs am gear carb      name
                                                                                               1  21   6  160 110  3.9 2.62 16.46  0  1    4    4 Mazda RX4
                                                                                               
                                                                                               require(lazyeval)
                                                                                               trend_fun <- function(data, org){
                                                                                               filter_criteria <- interp(~ colmn == org, colmn =as.name("name"))
                                                                                               z <- data %>% filter_(filter_criteria)
                                                                                               }
                                                                                               
                                                                                               trend_fun(mtcars, "Mazda RX4")
                                                                                               just_2 <- trend_fun(mtcars, "Mazda RX4")
                                                                                               just_2
                                                                                               
                                                                                               mpg cyl disp  hp drat   wt  qsec vs am gear carb      name
                                                                                               1  21   6  160 110  3.9 2.62 16.46  0  1    4    4 Mazda RX4
                                                                                               ```
                                                                                               
                                                                                               Adapted from: http://stackoverflow.com/questions/26492280/non-standard-evaluation-nse-in-dplyrs-filter-pulling-data-from-mysql
                                                                                               
                                                                                               ## Using `select()` in a function
                                                                                               
                                                                                               ```r
                                                                                               require(dplyr)
                                                                                               require(lazyeval)
                                                                                               data(mtcars)
                                                                                               head(mtcars)
                                                                                               
                                                                                               select_a_column <- function(data, column){
                                                                                               z <- data %>% select_("mpg", column)
                                                                                               return(z)
                                                                                               }
                                                                                               
                                                                                               select_a_column(mtcars, "hp")
                                                                                               ```
                                                                                               
                                                                                               ## Arrange in a function
                                                                                               
                                                                                               ```r
                                                                                               arrange_a_column <- function(data, column){
                                                                                               z <- data %>% arrange_(lazyeval::interp(~var, var = as.name(column)))
                                                                                               }
                                                                                               arrange_a_column(dat, "my_col_name")
                                                                                               ```
                                                                                               
                                                                                               ## Hypothesis tests and dplyr
                                                                                               
                                                                                               Technically possible, see broom and dplyr. 
                                                                                               I haven't (yet) managed this for a `chisq.test()`, but: 
                                                                                               
                                                                                               ```r
                                                                                               test <- as.data.frame(structure(list(organism.species.name = c("ec", "ec", "kp"), 
                                                                                               abx = c("ceph", "carb", "ceph"), 
                                                                                               n_sus_start = c(5L, 5L, 10L), 
                                                                                               n_res_start = c(10L, 5L, 5L), 
                                                                                               n_sus_end = c(10L, 5L, 5L), 
                                                                                               n_res_end = c(5L, 5L, 10L)), 
                                                                                               .Names = c("organism.species.name", "abx", "n_sus_start", "n_res_start", "n_sus_end", "n_res_end"), 
                                                                                               class = "data.frame", row.names = c(NA, -3L)))
                                                                                               
                                                                                               test %>% group_by(organism.species.name, abx) %>% 
                                                                                               mutate(p.val = chisq.test(
                                                                                               matrix(c(n_sus_start, n_res_start, n_sus_end, n_res_end), nrow = 2)
                                                                                               )$p.value
                                                                                               )
                                                                                               Source: local data frame [3 x 7]
                                                                                               Groups: organism.species.name, abx
                                                                                               
                                                                                               organism.species.name  abx n_sus_start n_res_start n_sus_end n_res_end    p.val
                                                                                               1                    ec ceph           5          10        10         5 0.144127
                                                                                               2                    ec carb           5           5         5         5 1.000000
                                                                                               3                    kp ceph          10           5         5        10 0.144127
                                                                                               ```
                                                                                               
                                                                                               ## tidyr
                                                                                               
                                                                                               ### wide to long
                                                                                               
                                                                                               ```r
                                                                                               stocks <- data.frame(
                                                                                               time = as.Date('2009-01-01') + 0:9,
                                                                                               X = rnorm(10, 0, 1),
                                                                                               Y = rnorm(10, 0, 2),
                                                                                               Z = rnorm(10, 0, 4)
                                                                                               )
                                                                                               stocks
                                                                                               time          X          Y          Z
                                                                                               1  2009-01-01  0.7877668 -0.2938621  2.6682693
                                                                                               2  2009-01-02  1.6313870  1.4439570 -2.9431324
                                                                                               3  2009-01-03 -1.2691361  0.4672836  3.6694435
                                                                                               4  2009-01-04 -1.2445373  0.9238149  1.8050772
                                                                                               5  2009-01-05  0.4251908  2.8406133  0.6584698
                                                                                               6  2009-01-06  1.4262145  3.5163632 -4.0324487
                                                                                               7  2009-01-07 -0.3578377 -1.0957885 -3.3847781
                                                                                               8  2009-01-08 -0.1034854 -4.8278236 -1.6016573
                                                                                               9  2009-01-09  0.8060879 -2.3204739 -2.2709526
                                                                                               10 2009-01-10 -1.9365305 -5.7522708  3.1133933
                                                                                               
                                                                                               gather(stocks, stock, price, -time)
                                                                                               time stock      price
                                                                                               1  2009-01-01     X  0.7877668
                                                                                               2  2009-01-02     X  1.6313870
                                                                                               3  2009-01-03     X -1.2691361
                                                                                               4  2009-01-04     X -1.2445373
                                                                                               5  2009-01-05     X  0.4251908
                                                                                               6  2009-01-06     X  1.4262145
                                                                                               ```
                                                                                               
                                                                                               ### long to wide
                                                                                               
                                                                                               `spread()` takes long data and makes it wide. 
                                                                                               One specifies the `key` which contains the cells that will become column headers and then `value` column which contains the data that will go in the cells beneath the new columns.
                                                                                               
                                                                                               ```r
                                                                                               data(infert)
                                                                                               long <- infert %>% 
                                                                                               mutate(id = seq_along(education)) %>% # create integer id
                                                                                               gather(key = variable, value = val, -id)
                                                                                               
                                                                                               head(long)
                                                                                               
                                                                                               # id  variable     val
                                                                                               #  1 education  0-5yrs
                                                                                               #  2 education  0-5yrs
                                                                                               #  3 education  0-5yrs
                                                                                               #  4 education  0-5yrs
                                                                                               #  5 education 6-11yrs
                                                                                               #  6 education 6-11yrs
                                                                                               
                                                                                               long %>% spread(key = variable, value = val) %>% 
                                                                                               head()
                                                                                               
                                                                                               # id age case education induced parity pooled.stratum spontaneous stratum
                                                                                               #  1  26    1    0-5yrs       1      6              3           2       1
                                                                                               #  2  42    1    0-5yrs       1      1              1           0       2
                                                                                               #  3  39    1    0-5yrs       2      6              4           0       3
                                                                                               #  4  34    1    0-5yrs       2      4              2           0       4
                                                                                               #  5  35    1   6-11yrs       1      3             32           1       5
                                                                                               #  6  36    1   6-11yrs       2      4             36           1       6
                                                                                               
                                                                                               ```
                                                                                               
                                                                                               For a more complex approach see [this SO post](http://stackoverflow.com/questions/30592094/r-spreading-multiple-columns-with-tidyr#30592293).
                                                                                               
                                                                                               See also `unite()` in tidyr
                                                                                               
                                                                                               If `spread()` tells you `"Error: All columns must be named"`, then you have `NA` values in your key column. 
                                                                                               
                                                                                               ```r
                                                                                               dat <- data_frame(
                                                                                               Person = rep(c("greg", "sally", "sue"), each=2),
                                                                                               Time = rep(c("Pre", "Post"), 3),
                                                                                               Score1 = round(rnorm(6, mean = 80, sd=4), 0),
                                                                                               Score2 = round(jitter(Score1, 15), 0),
                                                                                               Score3 = 5 + (Score1 + Score2)/2
                                                                                               )
                                                                                               
                                                                                               head(dat)
                                                                                               Source: local data frame [6 x 5]
                                                                                               
                                                                                               Person Time Score1 Score2 Score3
                                                                                               1   greg  Pre     84     85   89.5
                                                                                               2   greg Post     77     76   81.5
                                                                                               3  sally  Pre     74     73   78.5
                                                                                               4  sally Post     78     80   84.0
                                                                                               5    sue  Pre     81     81   86.0
                                                                                               6    sue Post     88     90   94.0
                                                                                               
                                                                                               dat %>% 
                                                                                               gather(temp, score, starts_with("Score")) %>% 
                                                                                               unite(temp1, Time, temp, sep = ".") %>% 
                                                                                               spread(temp1, score)
                                                                                               Source: local data frame [3 x 7]
                                                                                               
                                                                                               Person Post.Score1 Post.Score2 Post.Score3 Pre.Score1 Pre.Score2 Pre.Score3
                                                                                               1   greg          77          76        81.5         84         85       89.5
                                                                                               2  sally          78          80        84.0         74         73       78.5
                                                                                               3    sue          88          90        94.0         81         81       86.0
                                                                                               ```
                                                                                               
                                                                                               ## Joining
                                                                                               
                                                                                               Very similar to plyr, but the function name specifies the type of join rather than specifying within the function. 
                                                                                               
                                                                                               ```r
                                                                                               data3 <- left_join(data1, data2, by = "common.var")
                                                                                               ```
                                                                                               See SQL Venn for reminders. 
                                                                                               
                                                                                               Usefully, one can join by more than one variable. 
                                                                                               This is useful when a single variable does not provide a unique identifier. 
                                                                                               
                                                                                               ```r
                                                                                               x <- data.frame(var1 = c(1,1,2), var2 = c("a", "b", "b"), var3 = c("x", "y", "z"))
                                                                                               y <- data.frame(var1 = c(1,1,2), var2 = c("a", "b", "b"), var4 = c(20, 21, 22))
                                                                                               x
                                                                                               var1 var2 var3
                                                                                               1    1    a    x
                                                                                               2    1    b    y
                                                                                               3    2    b    z
                                                                                               y
                                                                                               var1 var2 var4
                                                                                               1    1    a   20
                                                                                               2    1    b   21
                                                                                               3    2    b   22
                                                                                               
                                                                                               z <- left_join(x, y, by = c("var1", "var2"))
                                                                                               z
                                                                                               var1 var2 var3 var4
                                                                                               1    1    a    x   20
                                                                                               2    1    b    y   21
                                                                                               3    2    b    z   22
                                                                                               ```
                                                                                               
                                                                                               ### anti_join
                                                                                               
                                                                                               Anti join retains those records which are in the left table but not the right and retains the columns of the left table. 
                                                                                               This is probably more useful than one might at first think. 
                                                                                               
                                                                                               ```
                                                                                               require(dplyr)
                                                                                               
                                                                                               # want those in 2, that do not appear in 1
                                                                                               
                                                                                               anti1 <- data.frame(id = c(1,1,1,2,3), abx = c("pen", "amx", "flu", "pen", "amx"))
                                                                                               anti2 <- data.frame(id = c(1,1,1,2,4), abx = c("pen", "amx", "flu", "flu", "pen"), 
                                                                                               res = c("r", "i", "s", "i", "s"))
                                                                                               anti1
                                                                                               id abx
                                                                                               1  1 pen
                                                                                               2  1 amx
                                                                                               3  1 flu
                                                                                               4  2 pen
                                                                                               5  3 amx
                                                                                               
                                                                                               anti2
                                                                                               id abx res
                                                                                               1  1 pen   r
                                                                                               2  1 amx   i
                                                                                               3  1 flu   s
                                                                                               4  2 flu   i
                                                                                               5  4 pen   s
                                                                                               
                                                                                               anti3 <- anti_join(anti2, anti1, by = "id")
                                                                                               anti3
                                                                                               id abx res
                                                                                               1  4 pen   s
                                                                                               
                                                                                               anti3 <- anti_join(anti2, anti1, by = c("id", "abx"))
                                                                                               anti3
                                                                                               id abx res
                                                                                               1  2 flu   i
                                                                                               2  4 pen   s
                                                                                               ```
                                                                                               