---
output:
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dates

```{r}
x <- "02/10/2017"
x
x <- as.Date(x, format = "%d/%m/%Y")
x
```

Date components:
Day

 * `%a` Abbreviated day of week
 * `%A` Full day of week
 * `%d` Numeric day of the month (01-31)
 * `%e` Numeric day of month with leading space for single digits
 * `%u` Numeric day of week, Monday = 1
 
Month 

 * `%b` Abbreviated month
 * `%B` Full month name
 * `%m` Numeric month (01-12)

Year

 * `%y` Year without century
 * `%Y` Year with century
 
So one can format a date:

```{r}
format(x, "%a")

format(x, "%A, %d %B %Y")
```

## Logical operators

```{r logic-operators}
x <- 5
x == 5
x < 1
x > 2
x < 5
x <= 5
x %in% c(1, 2, 3, 4, 5)
```

## Data types

### Data frames

These are the bread-and-butter data types where you'll do most of your work. 
They are rectangular data where each row forms an observation and each column is a variable. 
Technically, a dataframe is a `list` of equal-length `vectors`. 

### Making a dataframe

```{r making-a-dataframe}
x <- data.frame(id = c(1,2,3,4,5),
                x_var = c(4,2,9,4,7), 
                height = c("Tall", "Short", "Medium", "Tall", "Short"), 
                stringsAsFactors = FALSE)
x
```

### Accessing elements of a dataframe

Square-bracket indexing is used to access elements of a dataframe


```{r square-brckts}
# First row
x[1, ]

# First column
x[ , 1]

# Second row, third col
x[2, 3]

# all cols for 'short' observations
x[x$height == "Short", ]
```
## Missing values
