---
output: 
  html_document: 
    keep_md: true
    exclude: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

These are some really basic notes on spatial statistics, made as I read Applied Spatial Data Analysis with R, by R. S. Bivand *et al*.

## General approach

 1. Create a neighbour list
 1. Create neighbour weights
 1. Investigate spatial autocorrelation
 1. Investigate heterogeneity
 1. Model for clusters
 1. Map clusters
 
## Examining the data

 * `Z` log-trandformed rate
 * `PCTOWNHOME` percent population owning own home
 * `PCTAGE65P` percent population age >/= 65
 * `PEXPOSURE` inverse distance to closest TCE site (Trichloroethane)

```{r load-in-data}
# stupid, stupid, stupid win10
library(sp, lib.loc = "C:/Users/simon/OneDrive/Documents/R/win-library/3.3")
library(rgdal, lib.loc = "C:/Users/simon/OneDrive/Documents/R/win-library/3.3")
library(spdep, lib.loc = "C:/Users/simon/OneDrive/Documents/R/win-library/3.3")
library(DCluster, lib.loc = "C:/Users/simon/OneDrive/Documents/R/win-library/3.3")

# data from http://asdar-book.org/datasets.php?dataset=15
ny8 <- readOGR(".", "NY8_utm18")
head(ny8, n = 1)
# read in neighbour file. 
ny_nb <- read.gal("NY_nb.gal", region.id = row.names(as(ny8, "data.frame")))
class(ny_nb)
head(ny_nb)
```
 
The New York data has 281 census tract observations, covering both sparsely populated rural areas and densely populated urban areas.
Cases of leukaemia are recorded by tract. 
 
## Creating a neighbour list

Creates a matrix with one row for each area in data set. 
Demonstrated on a subset of the NY data.
These are contiguity neighbours. 
Areas with adjoining borders are neighbours. 
Other methods are available, such as distance.

```{r neighbour-list}
syr <- ny8[ny8$AREANAME == "Syracuse city", ]
sy_nb <- poly2nb(syr)
coords <- coordinates(syr) # just for plotting
plot(syr)
plot(sy_nb, coords, add = TRUE)
rm(coords)
```

## Calculating spatial weights

Conversion style `W` by default. 
Weights for each area are standardised to sum to 1

```{r spatial-weights}
syr_wts <- nb2listw(sy_nb)
syr_wts
```

## Spatial autocorrelation
Using the complete New York data. 
Tests for spatial autocorrelation in number of cases by census tract.
Observe significant spatial autocorrelation, with neighbouring areas having similar values. 

```{r morans-I}
moran.test(ny8$Cases, listw = nb2listw(ny_nb))
```

## Simultaneous autoregressive models
Uses a regression on values from other areas to account for spatial dependence. 

```{r sar}
# spautolm() from spdep
ny_sar <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = ny8, 
                   listw = nb2listw(ny_nb, style = "B"))
summary(ny_sar)
```

Lambda = 0.04, p = 0.02 indicates significant spatial correlation in residuals. 

Alternatively, can weight by population of census tracts:
(not currently run as I don't know where POPS came frome)
```{r}
ny_lm <- lm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = ny8, 
                   weights = POP8)
summary(ny_lm)
```

Can also use population weights in a `spautolm` model
```{r}
ny_sar2 <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = ny8, 
                   listw = nb2listw(ny_nb, style = "B"), weights = POP8)
summary(ny_sar2)
```
Interesting results from this. 
There is no longer spatial autocorrelation, and proximity to TCE site is now significant.
Lower AIC values indicate better model fits. 

## Conditional autoregressive models
Conditional distribution of spatial error terms. 
Only spatial errors of neighbours are used, rather than all spatial errors. 

```{r}
ny_car <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = ny8, 
                   listw = nb2listw(ny_nb, style = "B"), weights = POP8, 
                   family = "CAR")
summary(ny_car)
```


## Calculating expected cases
 1. Calculate the overall rate
 1. Multiply population at geographical unit by rate to get expected cases at geographical unit (p. 314)
 
## Detecting clusters of disease

```{r new-data}
rm(list = ls())
data("nc.sids")
ls()
class(nc.sids)
head(nc.sids)
```

Functions used later in this process have absolute requirements for two columns:
`Observed` and `Expected`. 

This next process calculates these values
```{r}
nc.sids$Observed <- nc.sids$SID74
nc.sids$Population <- nc.sids$BIR74
r <- sum(nc.sids$Observed) / sum(nc.sids$Population)
nc.sids$Expected <- nc.sids$Population * r
nc.sids$smr <- nc.sids$Observed/nc.sids$Expected
```

 1. First test homogeneity of relative risks
  * Chisq test for differences between expected and observed values as above. `DCluster::achisq.test()`

Requires a data frame for input. 
If you are using a spatialpoly, you can use `as(poly_name, "data.frame")
or `poly_name@data`

```{r homogeneity}
achisq.test(Observed ~ offset(log(Expected)),
            data = nc.sids, model = "multinom", 999)
```

There is significant evidence of overdispersion. 

```{r}
pottwhitt.test(Observed ~ offset(log(Expected)),
            data = nc.sids, model = "multinom", 999)
```

## Tango's test of global clustering

```{r}
coords <- cbind(nc.sids$x, nc.sids$y)
dlist <- dnearneigh(coords, 0, Inf)
dlist <- include.self(dlist)
dlist.d <- nbdists(dlist, coords = coords)
phi <- 100
col.W.tango <- nb2listw(dlist, 
                        glist = lapply(dlist.d, 
                                       function(x, phi){
                          exp(-x/phi)
                        }, phi = phi), style = "C")

tango.test(Observed ~ offset(log(Expected)),
           data = nc.sids, model = "negbin", 999, listw = col.W.tango, 
           zero.policy = TRUE)
```

## Locating clusters
Using Kulldorf's statistic
`calculate.mle()` requires `Expected` and `Observed` and data as a data.frame

```{r}
mle <- calculate.mle(nc.sids, model = "negbin")
the_grid <- nc.sids[, c("x", "y")] # another data.frame
knresults <- opgam(data = nc.sids, thegrid = the_grid, alpha = 0.05,
                   iscluster = kn.iscluster, 
                   fractpop = 0.15, # fraction of total population to use
                   R = 99, # Number of repetitions for bootstrapping
                   model = "negbin", mle = mle)
knresults
```

### Plot clusters

```{r plot}
#Plot centroids
plot(nc.sids$x, nc.sids$y, xlab="Easting", ylab="Northing")
#Plot points marked as clusters
points(knresults$x, knresults$y, col="red", pch="*")
```